# Bringing Everything Together. Part II {#session8}  

## Continuing with the exercise 10 and two-way ANOVA

Our dataframe should be all ready now. You might want to save it to a datafile.

We will be going through a number of steps to check ANOVA assumptions and follow-up with any necessary post-hoc analysis. As we do that we will also be generating some graphs which will give us the opportunity to discuss **ggplot2** and an extension of it called **ggpubr**. Our main package for the anova will be the **rstatix** package.

First, let's check that you completed **exercise 10**. 

```{r eval=FALSE, include=FALSE}
library(tidyverse)

stroop <- read_csv("stroopy stroop.csv")
stroop <- stroop %>% 
  mutate(RT = stroop_resp.rt*1000,
        group = case_when(
        participant < 2000 ~ "control",
        participant >= 2000 ~ "gambling"),
        salience = case_when(
          str_sub(Image,1,1) == 'G'~"Gambling",
          str_sub(Image,1,1) == 'N'~"Neutral")) %>% 
  filter(stroop_resp.rt > 0.3 & stroop_resp.rt < 3.0 
         & stroop_trials.thisTrialN > 2)
  
stroop2 <- stroop %>% 
  group_by(participant, group, salience) %>% 
  summarise(RT = mean(RT)) %>% 
  ungroup() 
```

We will now start working on our two-way analysis of variance. First we will load the necessary libraries and convert our IVs into factors.

```{r two-way-anova, eval=FALSE}
library(tidyverse)
library(ggpubr)
library(rstatix)

## make sure you have run the code above
## converting group and salience to factors
stroop2 <- stroop2 %>% 
  mutate(group = as.factor(group),
         salience = as.factor(salience))

## let's inspect our RT visually
ggboxplot(stroop2, x = "salience", y = "RT", color = "group")

##perhaps it would be more meaningful to also see boxplots per group
## we will also assign it to a variable so we can use it later
box1 <- ggboxplot(stroop2, x = "group", y = "RT", color = "salience")

```


We will now proceed with checks for outliers and anova assumptions. As this is a 2x2 some of the following checks are not meaningful but we will include them for reference.

```{r eval=FALSE}
## check for outliers

stroop2 %>%
  group_by(salience, group) %>%
  identify_outliers(RT)

## Normality check
stroop2 %>% 
  group_by(salience, group) %>%
  shapiro_test(RT)

## you may also want to check this visually
## using a QQplot

ggqqplot(stroop2, "RT", ggtheme = theme_bw()) +
  facet_grid(group ~ salience)

## Homogeneity of variance assumption
stroop2 %>% 
  group_by(salience) %>%
  levene_test(RT ~ group)

## the assumption of sphericity will be checked during anova

```

Assuming everything is in order we can proceed to carrying out the anova.

```{r eval=FALSE}
## anova time
# first describe the model
anova.model <- anova_test(
  data = stroop2, 
  dv = RT, 
  wid = participant,  ## this is our participant id variable
  between = group, 
  within = salience
)

# then get the anova table
get_anova_table(anova.model)

```

As we can see we have a significant interaction as well as a main effect.
We have to follow up the interaction. 

```{r eval=FALSE}
## First, We will need to regroup our data

simple.model <- stroop2 %>%
  group_by(group) %>%
  anova_test(dv = RT, wid = participant, within =  salience) %>%
  get_anova_table() %>%
  adjust_pvalue(method = "bonferroni")

simple.model


## or we could run pairwise comparisons
pair.model <- stroop2 %>%
  group_by(salience) %>%
  pairwise_t_test(RT ~ group, p.adjust.method = "bonferroni")

pair.model


## it would also be useful to report means and sd
stroop2 %>%
  group_by(salience, group) %>%
  get_summary_stats(RT, type = "mean_sd")
```

From here we can also generate plots if we want to as we already all the necessary stats figures.

```{r eval = FALSE}
pair.model <- pair.model %>% add_xy_position(x = "salience")


box1 + 
  stat_pvalue_manual(pair.model, tip.length = 0, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(anova.model, detailed = TRUE),
    caption = get_pwc_label(pair.model)
  )

## or with a barchart
box2 <- ggbarplot(stroop2, x = "group", y = "RT", width = 0.5,
                  color = "salience",
                  fill = "salience",
                  position = position_dodge(0.5),
                  add = "mean_se", add.params = list(color = "black"), 
                  label = TRUE,
                  lab.vjust = -2, lab.nb.digits = 2)
box2


```

As mentioned earlier **ggpubr** is a package that calls **ggplot2**. It is really versatile and can cover all our needs in terms of analysis and publications. However, ggplot2 offers many more features that you can try to explore on your own.

## Exercise 11

Now that you are familiar with the two-way anova you should attempt the three-way anova based on the data that we prepared in the previous session.
