---
title: "Comparing Means"
author: "Lazaros Gonidis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objectives  

By the end of today's session you should a good understanding on how to perform the following tasks using R:   

1. Independent samples parametric t-test (Student and Welch)   
2. Paired samples parametric t-test   
3. Non-parametric independent samples test (Mann-Whitney test)   
4. Non-parametric paired samples test (Wilcoxon test)   
5. Acquire effect-sizes    
6. Converting a dataframe from wide to long format and vice-versa   






## Independent samples parametric t-test (Student and Welch)   

When we are comparing means for independent samples we usually employ Student's t-test. However, in order to do so we have to check that our variable meets two assumptions:   
1. Our measurements are normally distributed   
2. Our measurements have equal variance   

If the first assumption is violated then we have to carry out a non-parametric equivalent (Mann-Whitney test).     
When our second assumption is violated we have to carry out a Welch t-test (still parametric). So, in practice before carrying out an indepedent samples t-test we have to check these two assumptions first. Let's see below how we do this. First, we load the datafile **stroopfinal.csv** into a dataframe.


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rstatix)

df <- read_csv("stroopfinal.csv")

head(df)

```




